---
title: "The free group in R"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Free group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE)
options(width = 80, tibble.width = Inf)
```

# The freegroup package

This package provides some functionality for manipulating the free
group on a finite list of symbols.

There is an excellent [wikipedia
page](https://en.wikipedia.org/wiki/Free_group).

Basically, the free group~$\left(X,\circ\right)$ on a
set $S=\{a,b,c,...,z\}$ is the set $X$ of *words* that are things like
$c^{-4}bb^2aa^{-1}ca$.  Usually one works only with words that are in
*reduced form*, which has successive powers of the same symbol
combined, in this case it would be $c^{-4}b^3ca$; see how the $a$ term
in the middle has vanished.

The group operation is juxtaposition; thus, for example
$a^2b^{-3}c^2\circ c^{-2}ba =a^2b^{-3}c^2c^-2ba =a^2b^{-2}ba$.

A word is represented by a two-row integer matrix: The top row is the
integer representation of the symbol and the second row is the
corresponding power.  For example, say we want to
represent~$a^2b^{-3}ac^2a^{-2}$ we would identify $a$ as 1, $b$ as 2, etc
and write


```{r}
(M <- rbind(c(1,2,3,3,1),c(2,-3,2,3,-2)))
```

(see how negative entries in the second row correspond to negative
powers).  Then to convert to a more useful form we would have

```{r}
library(freegroup)
(x <- free(M))
```

The representation for R object `x` is still a two-row matrix, but the
print method is active and uses a more visually appealing scheme.

We can coerce strings to free objects:


```{r}
(y <- as.free("aabbbcccc"))
```

The free group operation is simply juxtaposition, represented here by
the plus symbol, "`+`":


```{r}
x+y
```

See how the $a$ "cancels out" in the juxtaposition.  Note that
concatenation is associative.  Multiplication is also defined.
Suppose we want to concatenate 5 copies of `x`:

```{r}
x*5
```

The package is vectorized:

```{r}
x*(0:3)
```

There are a few methods for creating free objects:

```{r}
abc(1:9)
rfree(10,4)
```

Inverses are calculated using unary or binary minus:

```{r}
(u <- rfree(10,4))
-u
u-u
```


We can take the "sum" of a vector of free objects simply by
juxtaposing the elements:

```{r}
sum(u)
```

Powers are defined as per group conjugation: `x^y == y^{-1}xy` (or,
written in additive notation, `-y+x+y`):

```{r}
u
z <- alpha(26)
u^z
```


Thus:

```{r}
sum(u^z) == sum(u^z)
```




If we have more than 26 symbols the print method runs out of letters:

```{r}
free(rbind(1:30,1))
```

If this is a problem (it might not be: the print method might not be
important) it is possible to override the default symbol set:

```{r}
options(symbols = state.abb)
free(rbind(1:50,1))
```
