---
title: "The free group in R"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Free group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE)
options(width = 80, tibble.width = Inf)
```

# The freegroup package

This package provides some functionality for manipulating the free
group on a finite list of symbols.

There is an excellent [wikipedia
page](https://en.wikipedia.org/wiki/Free_group).

Informally, the *free group* $\left(X,\circ\right)$ on a set
$S=\{a,b,c,...,z\}$ is the set $X$ of _words_ that are objects like
$W=c^{-4}bb^2aa^{-1}ca$, with a group operation of string
juxtaposition.  Usually one works only with words that are in *reduced
form*, which has successive powers of the same symbol combined, so $W$
would be equal to $c^{-4}b^3ca$; see how $b$ appears to the third
power and the $a$ term in the middle has vanished.

The group operation of juxtaposition is formally indicated by $\circ$,
but this is often omitted in algebraic notation; thus, for example
$a^2b^{-3}c^2\circ c^{-2}ba =a^2b^{-3}c^2c{^-2}ba =a^2b^{-2}ba$.

In R package ```freegroup```, a word is represented by a two-row
integer matrix: The top row is the integer representation of the
symbol and the second row is the corresponding power.  For example,
say we want to represent $a^2b^{-3}ac^2a^{-2}$ we would identify $a$
as 1, $b$ as 2, etc and write


```{r}
(M <- rbind(c(1,2,3,3,1),c(2,-3,2,3,-2)))
```

(see how negative entries in the second row correspond to negative
powers).  Then to convert to a more useful form we would have

```{r}
library("freegroup")
(x <- free(M))
```

The representation for R object `x` is still a two-row matrix, but the
print method is active and uses a more visually appealing scheme.

We can coerce strings to free objects:


```{r}
(y <- as.free("aabbbcccc"))
```

The free group operation is simply juxtaposition, represented here by
the plus symbol, "`+`":


```{r}
x+y
```

(see how the $a$ "cancels out" in the juxtaposition).  One motivation
for the use of "`+`" rather than  "`*`" is that Python uses ```+```
for appending strings:

```
>>> "a" + "abc"
'aabc'
>>> 
```

However, note that the ```+``` symbol is usually reserved for
commutative operations; note that string juxtaposition is associative.
Multiplication by integers---denoted in ```freegroup``` idiom by
```*```---is also defined.  Suppose we want to concatenate 5 copies of
`x`:

```{r}
x*5
```

The package is vectorized:

```{r}
x*(0:3)
```

There are a few methods for creating free objects, for example:

```{r}
abc(1:9)
```

And we can also generate random free objects:

```{r}
rfree(10,4)
```

Inverses are calculated using unary or binary minus:

```{r}
(u <- rfree(10,4))
-u
u-u
```


We can take the "sum" of a vector of free objects simply by
juxtaposing the elements:

```{r}
sum(u)
```

Powers are defined as per group conjugation: `x^y == y^{-1}xy` (or,
written in additive notation, `-y+x+y`):

```{r}
u
z <- alpha(26)
u^z
```


Thus:

```{r}
sum(u^z) == sum(u^z)
```




If we have more than 26 symbols the print method runs out of letters:

```{r}
free(rbind(1:30,1))
```

If this is a problem (it might not be: the print method might not be
important) it is possible to override the default symbol set:

```{r}
options(symbols = state.abb)
free(rbind(1:50,1))
```
